Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-01-08 response-to-reviews.Rmd

l258 under soil wet conditions?

*doing this because I don't think we want to talk about soil conditions. I've always had atmospheric conditions in my mind as I've thought about this.*

L298-299 Far red to red light: I think this could be developed here or somewhere else. You covered the vertical variation in light quantity inside the canopy, but you could also report the vertical change in light quality (light spectrum) as the leaves absorb preferentially some wavelengths but reflects or transmit others.

**Added the following lines:**
**Canopy foliage absorb a large portion of PAR, and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers. Along with decreasing PAR, this entails a decrease in red to far red ratio of light as it reaches the forest floor, where understories are enriched in near infra red radiation and absorb light in wavelengths of 700-1000nm [de Castro 2000; @poorterLeafOpticalProperties2000]**

@@

**Done**

line52: 'with decreases being more commonly documented across the world’s forests'

line128: '… but understory leaf area density is often relatively high in the understory as well' sounds redundant. Please reword.

*removed the second 'understory' word here:* 
*'… but understory leaf area density is often relatively high in the understory as well' sounds redundant.*

@@

line140 'Light … decreases from the canopy top to the forest floor' – why isn’t foliage clumping mentioned here amongst the other factors?

**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor. The profile shape of this light is modified by leaf area density, leaf clumping, canopy height, and canopy structure across species and forest types (Fig. 2a-d, Supporting Information Figure S1, Koike et al., 2001).**

### 2022-01-08 vertical-thermal-review.Rmd
updates

Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).
Leaf area density (*i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Fig. `r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`). 
Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers, but understory leaf area density is often relatively high in the understory as well, sometimes with undulating patterns [e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a,@terborghVerticalComponentPlant1985; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989]. 
In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be predominantly accumulated in the lower canopy or understory layers [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`a, @baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001;  @bonan_ecological_2016]. 
Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @hanberryOpenForestEcosystems2020a]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019]. 
In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.

>>>>

Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).
Leaf area density (*i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Fig. `r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`). 
Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers, but understory leaf area density is often relatively high as well, sometimes with undulating patterns [e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a,@terborghVerticalComponentPlant1985; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989]. 
In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be predominantly accumulated in the lower canopy or understory layers [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`a, @baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001;  @bonan_ecological_2016]. 
Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @hanberryOpenForestEcosystems2020a]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019]. 
In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.
